---
title: "Discnorm: Testing  for underlying normality in ordinal datasets"
author: "Njål Foldnes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
citation_package: natbib
bibliography: bibsel.bib
biblio-style: apalike
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(discnorm )
```

The discnorm package uses bootstrapping to help determine whether the commonly assumed normality assumption is tenable for an ordinal dataset. Researchers wanting to proceed with ordinal SEM based on polychoric correlations need first to check that the normality assumption is not violated. 

The procedure is fully described in @fold1

## Examples

The main procedure is bootTest() which operates on an ordinal dataset and returns a p-value associated with the null-hypothesis of underlying normality.  Let us first use the test for a dataset that is produced by underlying normality.
```{r}
#let us discretize an underlying normal vector
# with moderate correlation 
rho <- 0.3
Sigma <- diag(5)
Sigma[Sigma !=1] <- rho
set.seed(1234)
norm.sample  <- MASS::mvrnorm(n=200, mu=rep(0,5), Sigma=Sigma)
# let us discretize into 4 categories
disc.sample <- discnorm::getdisc(norm.sample, rep(list(c(-Inf, -1, 1, 2, Inf)),5))

#check for underlying normality
bootTest(disc.sample)
# we have no evidence against the null hypothesis of underlying normality
```
And let us discretize a non-normal dataset

```{r}
nonnorm.sample <- data.frame(norm.sample[, 1:4], norm.sample[,1]*norm.sample[,2])
disc.sample2 <- discnorm::getdisc(nonnorm.sample, rep(list(c(-Inf, -1, 1, 2, Inf)),5))
bootTest(disc.sample2)
# rejected!
```


## References
