---
title: "Discnorm: Testing  for underlying normality in ordinal datasets"
author: "Njål Foldnes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
citation_package: natbib
bibliography: bibsel.bib
biblio-style: apalike
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(discnorm )
```

The discnorm package uses bootstrapping to help determine whether the commonly assumed normality assumption is tenable for an ordinal dataset. Researchers wanting to proceed with ordinal SEM based on polychoric correlations need first to check that the normality assumption is not violated. 

The procedure is fully described in @fold1

## An example

The main procedure is bootTest() which operates on an ordinal dataset and returns a p-value associated with the null-hypothesis of underlying normality. 
```{r}
library(psych)
library(dplyr)

data(bfi)# from psych package
set.seed(1234)
#select a rather small subsample from the big-five data
my.sample <- bfi[sample(nrow(bfi), size=200 , replace=T),]
#check for underlying normality in the agreeableness indicators
bootTest(my.sample[, paste0("A", 1:5)])
# we have some evidence against the null hypothesis of underlying normality
#is there underlying normality for agreeableness and openness combined?
bootTest(my.sample[, c(paste0("A", 1:5), paste0("O", 1:5))])
# strong evidence against underlying normality for these ten indicators
```


## References
